The Multi-Stencil Framework: A Component-Based Approach to Orchestrate Stencils

### 2017-03-03 ARTICLES/IJPP/articleR1/article.tex
corrections + evaluation en cours
abstract + titre

%----------------------------------------
\title{Orchestrating stencils with the Multi-Stencil Framework}
%----------------------------------------
\author{H\'el\`ene Coullon \and Julien Bigot \and Christian Perez}
%----------------------------------------
\institute{H\'el\`ene Coullon \and Christian Perez \at
              INRIA, LIP, ENS Lyon, 46 Allée d'Italie, 69364 Lyon Cedex 07, France\\
              \email{helene.coullon@inria.fr} \email{christian.perez@inria.fr}
           \and
           Julien Bigot \at
           Maison de la Simulation, CEA, CNRS, Univ. Paris-Sud, UVSQ, Universit\'e Paris-Saclay, 91191 Gif-sur-Yvette, France\\
           \email{julien.bigot@cea.fr}
}
%----------------------------------------
\maketitle
%----------------------------------------
\HC{change the title, improve it}
\HC{is it possible to have authors on a single line ?}
\begin{abstract}
\HC{has to be revised according to modifications}
As the computation power of modern high performance architectures increases, their heterogeneity and complexity also become more important. One of the big challenges of exascale is to get programming models which gives access to high performance computing (HPC) to many scientists and not only to a few HPC specialists. One relevant solution to ease parallel programming for scientists is Domain Specific Language (DSL). However, one problem to avoid with DSLs is to not design a new DSL each time a new domain or a new problem has to be solved. This phenomenon happens for stencil-based numerical simulations, for which a large number of languages has been proposed without code reuse between them. The Multi-Stencil Language (MSL) presented in this paper is a language common to any kind of mesh used into a stencil-based numerical simulation. It is said that MSL is mesh-agnostic. Actually, from the description of a numerical simulation, MSL produces an empty parallel pattern, or skeleton, of the simulation which will be filled using other existing parallel languages and libraries. Thus, MSL, by finding a common language for different kinds of stencil-based simulation, facilitates code reuse. MSL is evaluated on a real case simulation which solves shallow-water equations. It is shown that MSL does not introduce overheads on data parallelism up to 16.384 cores, and that the hybrid parallelism (data and task) introduced improves performance of the simulation.

\keywords{Domain Specific Language (DSL) \and stencil \and numerical simulation \and mesh-agnostic \and data parallelism \and task parallelism \and scheduling \and MPI \and OpenMP}
\end{abstract}
%------------------------------------------------------------------------------
\section{Introduction}
\label{sect:introduction}
\input{introduction.tex}
%------------------------------------------------------------------------------
\section{The Component-Based Multi-Stencil Framework}
\label{sect:msf}
\input{msf.tex}
%------------------------------------------------------------------------------
\section{Meta-modelization of a Multi-Stencil Program}
\label{sect:formalism}
\input{formalism.tex}
%------------------------------------------------------------------------------
\section{Generic Asssembly and The Multi-Stencil Language}
\label{sect:msl}
\input{msl.tex}
%------------------------------------------------------------------------------
\section{The Multi-Stencil Compiler}
\label{sect:parallelism}
\input{parallelism.tex}
%------------------------------------------------------------------------------
\section{Static Scheduling and performance model}
\label{sect:msp}
\input{msp.tex}
%------------------------------------------------------------------------------
\section{Implementation}
\label{sect:comp}
\input{compiler.tex}
%------------------------------------------------------------------------------
\section{Evaluation}
\label{sect:eval}
\input{evaluation.tex}
%------------------------------------------------------------------------------
\section{Related work}
\label{sect:rel}
\input{related.tex}
%------------------------------------------------------------------------------
\section{Conclusion}
\label{sect:concl}
\input{conclusion.tex}
%------------------------------------------------------------------------------
\begin{acknowledgements}
This work has partially been supported by the PIA ELCI project of the French FSN.
This  work  was  granted  access  to  the  HPC  resources  of TGCC under  the 
allocations t2015067470, x2016067617 and AP010610191 made by GENCI.
\end{acknowledgements}
%----------------------------------------
\bibliographystyle{plain}
\bibliography{biblio}
%----------------------------------------
\end{document}
%----------------------------------------

>>>>

%----------------------------------------
\title{The Multi-Stencil Framework: A Component-Based Approach to Orchestrate Stencils}
%----------------------------------------
\author{H\'el\`ene Coullon \and Julien Bigot \and Christian Perez}
%----------------------------------------
\institute{H\'el\`ene Coullon \at
              DAPI IMT Atlantique, LS2N, Inria. Nantes, France\\
              \email{helene.coullon@inria.fr}
	\and
           Julien Bigot \at
           Maison de la Simulation, CEA, CNRS, Univ. Paris-Sud, UVSQ, Universit\'e Paris-Saclay, 91191 Gif-sur-Yvette, France\\
           \email{julien.bigot@cea.fr}           
           \and
	Christian Perez \at
              Univ. Lyon, Inria, CNRS, ENS de Lyon. Lyon, France\\
              \email{christian.perez@inria.fr}
}
%----------------------------------------
\maketitle
%----------------------------------------
\begin{abstract}
As the computation power of modern high performance architectures increases, their heterogeneity and complexity also become more important. One of the big challenges of exascale is to reach programming models that give access to high performance computing (HPC) to many scientists and not only to a few HPC specialists. One relevant solution to ease parallel programming for scientists is Domain Specific Language (DSL). However, one problem to avoid with DSLs is to mutualize existing codes and libraries instead of implemented from scratch. For example, this phenomenon occurs for stencil-based numerical simulations, for which a large number of languages has been proposed without code reuse between them. 
The Multi-Stencil Framework (MSF) presented in this paper combines a meta-model to component-based programming models to enhance code reuse and separation of concerns in the specific case of stencils. MSF can easily choose one parallelization technique or another as well as one optimization or another, as well as one back-end implementation or another. It is shown that MSF can reach same performances than a non-component-based MPI implementation over 16.384 cores. Finally, the performance model of the framework for hybrid parallelization is validated by evaluations.

\keywords{Component programming models \and Domain Specific Language (DSL) \and Stencil \and Numerical simulation \and Data parallelism \and Task parallelism \and Scheduling \and MPI \and OpenMP}
\end{abstract}
%------------------------------------------------------------------------------
\section{Introduction}
\label{sect:introduction}
\input{introduction.tex}
%------------------------------------------------------------------------------
\section{The Component-Based Multi-Stencil Framework}
\label{sect:msf}
\input{msf.tex}
%------------------------------------------------------------------------------
\section{Meta-modelization of a Multi-Stencil Program}
\label{sect:formalism}
\input{formalism.tex}
%------------------------------------------------------------------------------
\section{Generic Asssembly and The Multi-Stencil Language}
\label{sect:msl}
\input{msl.tex}
%------------------------------------------------------------------------------
\section{The Multi-Stencil Compiler}
\label{sect:parallelism}
\input{parallelism.tex}
%------------------------------------------------------------------------------
%\section{Static Scheduling and performance model}
%\label{sect:msp}
%\input{msp.tex}
%------------------------------------------------------------------------------
%\section{Implementation}
%\label{sect:comp}
%\input{compiler.tex}
%------------------------------------------------------------------------------
\section{Evaluation}
\label{sect:eval}
\input{evaluation.tex}
%------------------------------------------------------------------------------
\section{Related work}
\label{sect:rel}
\input{related.tex}
%------------------------------------------------------------------------------
\section{Conclusion}
\label{sect:concl}
\input{conclusion.tex}
%------------------------------------------------------------------------------
\begin{acknowledgements}
This work has partially been supported by the PIA ELCI project of the French FSN.
This  work  was  granted  access  to  the  HPC  resources  of TGCC under  the 
allocations t2015067470, x2016067617 and AP010610191 made by GENCI.
\end{acknowledgements}
%----------------------------------------
\bibliographystyle{plain}
\bibliography{article}
%----------------------------------------
\end{document}
%----------------------------------------

### 2017-03-03 ARTICLES/IJPP/articleR1/conclusion.tex
abstract + titre

In this paper, we have presented the Multi-Stencil Framework. MSF is built upon a meta-model of a multi-stencil program which has been presented in Section~\ref{sect:formalism}. From this meta-model has been designed, first, the generic component assembly of a multi-stencil program, and second, the domain specific language MSL that enables the description of a specific application by a numerician. From these entries, the MSF compiler, namely MSC, is able to generate automatically a parallel specific assembly of the simulation. This specific assembly represents the parallel orchestration of computations independently of implementation choices. Moreover, by combining a meta-model and component-based programming models MSF has the particularity to enhance separation of concerns and code-reuse of existing solutions (\eg SkelGIS and OpenMP in this paper). Finally, MSF has been evaluated on the real case simulation FullSWOF2D. Results have shown that even if component models have been used, strong and weak scalings of MSF (for its data parallelization technique) do not suffer from overheads. Results have also shown that the hybrid parallelization handled by MSF could increase performances when MPI communications are too heavy, which is new compared to existing solutions. Finally, we have presented evaluations on fusion optimization that demonstrates that fusion is not always a good choice, particularly when performing hybrid parallelism.

Many perspectives of future work raise from this work. First, even if using one distributed data structure or another is possible and facilitated by MSF, the HPC specialist still has to understand component programming models to add one. We think it could be interesting to also facilitate the work of the HPC specialist by proposing a semi-automated framework to add external tools to MSF.

@@

The same choice problem is raised by fusion evaluations of Section~\ref{sect:fus}. Actually, fusion optimization is not always relevant. Thus, for the same reasons than parallelization techniques, an actor algorithm could choose the best solution for a given execution. To design such actor algorithms different solutions should be studied, such as following an approximated behavior model, or using setup runs to use deep learning algorithms etc.

>>>>

In this paper, we have presented the Multi-Stencil Framework. MSF is built upon a meta-model of a multi-stencil program which has been presented in Section~\ref{sect:formalism}. From this meta-model has been designed, first, the generic component assembly of a multi-stencil program, and second, the domain specific language MSL that enables the description of a specific application by a numerician. From these entries, the MSF compiler, namely MSC, is able to generate automatically a parallel specific assembly of the simulation. This specific assembly represents the parallel orchestration of computations, independently of implementation choices. Moreover, by combining a meta-model and component-based programming models MSF has the particularity to enhance separation of concerns and code-reuse of existing solutions (\eg SkelGIS and OpenMP in this paper). Finally, MSF has been evaluated on the real case simulation FullSWOF2D. Results have shown that even if component models have been used, strong and weak scalings of MSF (for its data parallelization technique) do not suffer from overheads. Results have also shown that the hybrid parallelization handled by MSF could increase performances when MPI communications are too heavy, which is a new contribution compared to existing solutions. Finally, we have presented evaluations on fusion optimization that demonstrates that fusion is not always a good choice, particularly when performing hybrid parallelism.

Many perspectives of future work raise from this paper. First, even if using one distributed data structure or another is possible and facilitated by MSF, the HPC specialist still has to understand component programming models and the meta-model of the framework to add new back-ends. We think it could be interesting to also facilitate the work of the HPC specialist by proposing a semi-automated framework to add external tools to MSF.

@@

The same choice problem is raised by fusion evaluations of Section~\ref{sect:fus}. Actually, fusion optimization is not always relevant. Thus, for the same reasons than parallelization techniques, an actor algorithm could choose the best solution for a given execution. To design such actor algorithms different solutions should be studied, such as following an approximated behavior model, or using calibration runs to then use deep learning algorithms etc.

### 2017-03-03 ARTICLES/IJPP/articleR1/introduction.tex
abstract + titre

\HC{has to be revised according to modifications}

As the computation power of modern high performance architectures increases, their heterogeneity and complexity also become more important. For example, the current world's top supercomputer Tianhe-2~\footnote{\url{www.top500.org}} is composed of multi-cores processors and accelerators, and is able to reach a theoretical peak performance around thirty peta floating point operations par seconds. However, to be able to use such a machine, multiple programming models, such as MPI (Message Passing Interface), OpenMP, CUDA etc., and multiple optimization techniques, such as cache optimization, have to be combined. Moreover, current architectures evolution lets think that heterogeneity and complexity in HPC will continue to grow in future.

@@

One possible runtime execution model for HPC is to propose dynamic scheduling of task graphs combined to message passing models~\cite{Gautier:2013:XRS:2510661.2511383,Augonnet2011,wu:hal-01078359} (to be able to use more than one machine). Those models increase HPC code portability and reach an interesting performance onto heterogeneous architectures, which is interesting to reach exascale programming. At a higher abstraction level, general purpose parallel languages, such as OpenMP~\cite{660313} and OpenCL~\cite{Stone:2010:OPP:622179.1803953} follow the direction of task graph scheduling proposed by those execution models. However, for non-experts end-users, general purpose languages still are difficult to use and to tune for a given application onto a given architecture. The current easiest (with the higher abstraction level), but still efficient, programming model for end-users is Domain Specific Language (DSL) and domain specific frameworks. Such solutions are specific to the end-user domain and propose a grammar or an API which is easy to understand. Moreover, the compiler is able, because of the specific knowledge onto the targeted domain, to automatically apply parallelization and specific optimization to produce a high performance back-end code. Thus, a domain specific solutions are able to split end-user concerns from HPC concerns which is relevant to reach exascale programming models.

@@

We propose the Multi-Stencil Framework (MSF) that is built over a meta-model of multi-stencil simulations. MSF produces a parallel orchestration of a multi-stencil program without being aware of the underlying implementation choices (\eg distributed data structures, task scheduler etc.). Thanks to this meta-model MSF is able to easily switch from one parallelization technique to another and from one optimization to another. Moreover, as the meta-model is independent from implementation details, MSF can easily choose one back-end or another, thus easing code reuse of existing solutions. To ease composition of existing solutions, MSF uses Software Engineering capabilities of component-based programming models~\cite{Szyperski:2002:CSB:515228} where applications are defined as an assembly of building blocks, or components.

After a short overview of the Multi-Stencil Framework given in Section~\ref{sect:msf}, contributions of the paper are presented as follows: a meta-model of a multi-stencil program presented in Section~\ref{sect:formalism}; from this meta-model are built both a light and descriptive domain specific language, namely MSL, as well as a generic component assembly of the application described in Section~\ref{sect:msl}; the compiler MSC of the framework builds a parallel specialized version of the generic assembly from the application description with MSL, MSC and its performance model are described in Section~\ref{sect:parallelism}; finally a performance evaluation is detailed in Section~\ref{sect:eval}  and validates the performance model.

>>>>

As the computation power of modern high performance architectures increases, their heterogeneity and complexity also become more important. For example, the current world's top supercomputer Tianhe-2~\footnote{\url{www.top500.org}} is composed of multi-cores processors and accelerators, and is able to reach a theoretical peak performance around thirty peta floating point operations per second. However, to be able to use such machines, multiple programming models, such as MPI (Message Passing Interface), OpenMP, CUDA etc., and multiple optimization techniques, such as cache optimization, have to be combined. Moreover, current architectures evolution lets think that heterogeneity and complexity in HPC will continue to grow in future.

@@

One possible runtime execution model for HPC is to propose dynamic scheduling of task graphs combined to message passing models~\cite{Gautier:2013:XRS:2510661.2511383,Augonnet2011,wu:hal-01078359} (to be able to use more than one machine). These models increase HPC code portability and reach an interesting performance onto heterogeneous architectures, which is interesting to reach exascale programming. At a higher abstraction level, general purpose parallel languages, such as OpenMP~\cite{660313} and OpenCL~\cite{Stone:2010:OPP:622179.1803953} follow the direction of task graph scheduling proposed by those execution models. However, for non-experts end-users, general purpose languages still are difficult to use and to tune for a given application onto a given architecture. The current easiest (with the higher abstraction level), but still efficient, programming model for end-users is Domain Specific Language (DSL) and domain specific frameworks. Such solutions are specific to the end-user domain and propose a grammar or an API which is easy to understand. Moreover, the compiler is able, because of the specific knowledge onto the targeted domain, to automatically apply parallelization and specific optimization to produce a high performance back-end code. Thus, domain specific solutions are able to distinguish end-user concerns from HPC concerns which is relevant to open HPC to non specialists.

@@

We propose the Multi-Stencil Framework (MSF) that is built upon a meta-model of multi-stencil simulations. MSF produces a parallel orchestration of a multi-stencil program without being aware of the underlying implementation choices (\eg distributed data structures, task scheduler etc.). Thanks to this meta-model MSF is able to easily switch from one parallelization technique to another and from one optimization to another. Moreover, as the meta-model is independent from implementation details, MSF can easily choose one back-end or another, thus easing code reuse of existing solutions. To ease composition of existing solutions, MSF uses Software Engineering capabilities of component-based programming models~\cite{Szyperski:2002:CSB:515228}, where applications are defined as an assembly of building blocks, or components.

After a short overview of the Multi-Stencil Framework given in Section~\ref{sect:msf}, the paper is organized as follows. A meta-model of a multi-stencil program is presented in Section~\ref{sect:formalism}; from this meta-model are built both a light and descriptive domain specific language, namely MSL, as well as a generic component assembly of the application described in Section~\ref{sect:msl}; the compiler MSC of the framework is described in Section~\ref{sect:parallelism}; finally a performance evaluation is detailed in Section~\ref{sect:eval} .
